---
title: "Simulations Pre-Class Project"
date: "Due March 13, 2017 at 5:00pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 6)
```




#Project Goals:


With this project we will simulate a famoues probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    ```{r}
 money_broken=function(date,N=5000,start=1000,bet=100){
      stop_day=NA
      money=rep(NA,N)
      money[1]=start
      for(i in 2:(date+1)){
        money[i]=money[i-1]+(2*rbinom(1,1,0.5)-1)*bet
        if(money[i]<=0){
          return(TRUE)
        }
      } 
      return(FALSE)
    }
    prob=mean(replicate(5000,money_broken(100)))
    prob

    ```
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    ```{r}
    prob2=mean(replicate(5000,money_broken(500)))
    prob2
    ```
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
    ```{r}
      broken_date=function(N=5000,start=1000,bet=100){
      stop_day=NA
      money=rep(NA,N)
      money[1]=start
      for(i in 2:N){
        money[i]=money[i-1]+(2*rbinom(1,1,0.5)-1)*bet
        if(money[i]<=0){
          return(i-1)
        }
      }
      return(NA)
      }
    
    est_time=mean(replicate(5000,broken_date()),na.rm=TRUE)
    est_time
    ```
    d. the mean and variance of your bankroll after 100 hands (including busts).
    ```{r}
     bankroll=function(date,start=1000,bet=100){
      N=date+1
      stop_day=NA
      money=rep(NA,N)
      money[1]=start
      for(i in 2:N){
        money[i]=money[i-1]+(2*rbinom(1,1,0.5)-1)*bet
      if(money[i]<=0){
        return(0)
      }
      }
      return(money[N])
      }
    mean_100=mean(replicate(5000,bankroll(100)))
    var_100=var(replicate(5000,bankroll(100)))
    mean_100
    var_100
    ```
    e. the mean and variance of your bankroll after 500 hands (including busts).
    
```{r}
    mean_500=mean(replicate(5000,bankroll(500)))
    var_500=var(replicate(5000,bankroll(500)))
    mean_500
    var_500
    ```
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
    ```{r}
    wager_final=function(N,stat_rate,ori_wager,speed,max_rate=1,min_wager=0){
       N1=N+1
       wager=rep(NA,N1)
       rate=rep(NA,N1)
       wager[1]=ori_wager
       rate[1]=stat_rate
       for(i in 2:N1){
         measure=rbinom(1,1,rate[i-1])
         if(measure==1){
           rate[i]=rate[i-1]+speed
           if(rate[i]>max_rate){
             rate[i]=max_rate
           }
           wager[i]=wager[i-1]+1
         }else{
         rate[i]=stat_rate
          wager[i]=wager[i-1]-1
          if(wager[i]<min_wager){
            wager[i]=min_wager
          }
       }
       }
       final=wager[N1]
       return(final)
     }
    
    sample=replicate(100,wager_final(100000,0.48,100,0.01))
    exp=mean(sample)
    exp

    ## it is not fair since the result is much less than 100
    ```
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
    ```{r}
    for(i in 1:40){
      k=0.48+1/2000
      sample_new=replicate(100,wager_final(100000,k,100,0.01))
      print(k)
      exp_new=mean(sample_new)
      print(exp_new)
    }
     ##In this step we can see the starting probability should be in the range of 0.4873 and 0.4874
    for(j in 1:10){
      k=0.4873+j/100000
      sample_new=replicate(100,wager_final(100000,k,100,0.01))
      print(k)
      exp_new=mean(sample_new)
      print(exp_new)
    }
   ##Finally, it should be about 48.731%
    ```
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
    ```{r}
    for(i in 1:20){
      k=0.0105+i/10000
      sample_new=replicate(100,wager_final(100000,0.48,100,k))
      print(k)
      exp_new=mean(sample_new)
      print(exp_new)
     }
    ##In thIs step, we know the increment to a value should be in the range of 0.0108 and 0.0109
     for(i in 1:10){
      k=0.0108+1/100000
      sample_new=replicate(100,wager_final(100000,0.48,100,k))
      print(k)
      exp_new=mean(sample_new)
      print(exp_new)
     }
     ##Finally, it should be about 1.089%
