---
title: "Simulations In-Class Project"
date: "Due October 13, 2017 at 11:59pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:



With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 


Since you all have completed problem 1, you first step will be to work through each of your groupmates code for problem #1 and comment on what is happening. Then as a team move forward in on direction as you move on to the rest of the problems 2-5

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
 
```{r}

# This is a random walk of the gambler's ruin situation, using code shown 
# in the Markov Chains video of Statistical Programming in R Chapter 5
gamblers_ruin <- function(n, winprob=.5, start = 1000, bet = 100){
  output <- rep(NA, n)
  output[1] <- start
  for(i in 2:n){
    output[i] <- output[i-1] + 2*bet*rbinom(1, 1, winprob) - bet
  }
  return(output)
}

```

    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    
```{r}

# Modified gambler's ruin to return 1 if gambler has not busted and 0 if 
# gambler has busted by the nth hand
gamblers_ruin_bust <- function(n, winprob = .5, start = 1000, bet = 100){
  output <- rep(NA, n)
  output[1] <- start
  for(i in 2:n){
    output[i] <- output[i-1] + 2*bet*rbinom(1, 1, winprob) - bet
    if(output[i] == 0) break
  }
  if(output[i] == 0) return(0)
  else return(1)
  
}


# Replicate gives a vector of 0s and 1s of busts and non busts
# Simulate 5000 times and use it to approximate the probability of busting within 100 hands
prob_bust_100hands <- 1 - mean(replicate(5000, gamblers_ruin_bust(100)))
prob_bust_100hands
```

    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    
```{r}

# Use same function to simulate 5000 runs of 500 hands and use the result 
# to approximate the probability of busting within 500 hands
prob_bust_500hands <- 1 - mean(replicate(5000, gamblers_ruin_bust(500)))
prob_bust_500hands

```
    
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
    
```{r}

# Return which number hand the gambler busted and NA if the gambler has 
# not busted within 5000 hands
gamblers_ruin_bust_hand <- function(n, winprob = .5, start = 1000, bet = 100){
  output <- rep(NA, n)
  output[1] <- start
  for(i in 2:n){
    output[i] <- output[i-1] + 2*bet*rbinom(1, 1, winprob) - bet
    if(output[i] == 0) break
  }
  if(output[i] == 0) return(i)
  else return(NA)
  
}

# Calcluate the mean bust time
mean_bust_time <- mean(replicate(5000, gamblers_ruin_bust_hand(5000)), na.rm = TRUE)
mean_bust_time
```
    
    d. the mean and variance of your bankroll after 100 hands (including busts).
    
```{r}
# Modified the gambler's ruin to stop after busting (going to 0)
# Also, this function returns the final amount of money rather rather than
# the whole random walk
gamblers_ruin_stop_after_bust <- function(n, winprob=.5, start = 1000, bet = 100){
  output <- rep(NA, n)
  output[1] <- start
  for(i in 2:n){
    output[i] <- output[i-1] + 2*bet*rbinom(1, 1, winprob) - bet
    if(output[i] == 0) break
  }
  return(output[i])
}

# Simulate 100 hands and find mean, variance
hundred_hands <- replicate(5000, gamblers_ruin_stop_after_bust(100))
mean(hundred_hands, na.rm = TRUE)
var(hundred_hands, na.rm = TRUE)

```
    
    e. the mean and variance of your bankroll after 500 hands (including busts).
    
```{r}
# Modified the gambler's ruin to stop after busting (going to 0)
# Also, this function returns the final amount of money rather rather than
# the whole random walk
gamblers_ruin_stop_after_bust <- function(n, winprob=.5, start = 1000, bet = 100){
  output <- rep(NA, n)
  output[1] <- start
  for(i in 2:n){
    output[i] <- output[i-1] + 2*bet*rbinom(1, 1, winprob) - bet
    if(output[i] == 0) break
  }
  return(output[i])
}

# Simulate 500 hands and find mean, variance
fiveHundred_hands <- replicate(5000, gamblers_ruin_stop_after_bust(500))
mean(fiveHundred_hands, na.rm = TRUE)
var(fiveHundred_hands, na.rm = TRUE)


```
    
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.

    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    
```{r}
# Probability you have busted by 100th bet with winprob 18/38
prob_bust_100hands <- 1 - mean(replicate(5000, gamblers_ruin_bust(n=100, winprob=18/38)))
prob_bust_100hands
```

    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 

```{r}
# Probability you have busted by 500th bet with winprob 18/38
prob_bust_100hands <- 1 - mean(replicate(5000, gamblers_ruin_bust(n=500, winprob=18/38)))
prob_bust_100hands
```
    
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
    
```{r}
# Calcluate the mean bust time
mean_bust_time <- mean(replicate(5000, gamblers_ruin_bust_hand(n=5000, winprob=18/38)), na.rm = TRUE)
mean_bust_time
```
    
    d. the mean and variance of your bankroll after 100 hands (including busts).
 
```{r}    
# Simulate 100 hands and find mean, variance
hundred_hands <- replicate(5000, gamblers_ruin_stop_after_bust(n=100, winprob=18/38))
mean(hundred_hands, na.rm = TRUE)
var(hundred_hands, na.rm = TRUE)
```

    e. the mean and variance of your bankroll after 500 hands (including busts).
    
```{r}
# Simulate 500 hands and find mean, variance
fiveHundred_hands <- replicate(5000, gamblers_ruin_stop_after_bust(n=500, winprob=18/38))
mean(fiveHundred_hands, na.rm = TRUE)
var(fiveHundred_hands, na.rm = TRUE)
```
    

3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
    
```{r}

# This function simulates the described situation, adding a percent of
# success with every success and going back down to .48 with every failure
# We start at 0$ as we are assuming we cannot go bust
game_chain <- function(n, start = 0, bet = 100, winprob = .48, prob_inc = .01){
  output <- rep(NA, n)
  output[1] <- start
  # Have our win probability in our function environment
  winprob1 = winprob
  for(i in 2:n){
    output[i] <- output[i-1] + 2*bet*rbinom(1, 1, winprob1) - bet
    # Add one percent for every win
    if(output[i] > output[i-1] && winprob1 < 1){
      winprob1 <- winprob1 + prob_inc
      #print(winprob1)
      # Consider if the win prob becomes 1 or greater than 1, then automatically finish the loop
      if(winprob1 >= 1){
        for(j in (i+1):n){
        output[j] <- output[j-1] + bet
        }
        return(output[n])
        break
      }
    } 
    else {
      winprob1 <- winprob
      #print(winprob1)
    }
  }  
  # Return the final amount of money
  return(output[n])
}

# Calculate expected value
expected_value <- mean(replicate(100, game_chain(n=100000)))
expected_value

```
    
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
    
```{r}
# It looks like 49% or 48.9% is the fairest the game will be within 3 decimal places
expected_value <- mean(replicate(100, game_chain(n=100000, winprob = .49)))
expected_value

# This runs the game with starting probability at 48.9%
expected_value1 <- mean(replicate(100, game_chain(n=100000, winprob = .489)))
expected_value1
```

*This can be done automatically. We can test boundaries and if they yield answers on both sides of zero, we can test the midpoint of them. If the midpoint is greater than 0, we can test the midpoint of the original lower bound and the first midpoint and vice versa if the first midpoint is less than 0. We can continue this until we get close to 0 which indicates a fair game.*
    
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 

```{r}    
# Within 3 decimal places, a probability increment of 1.2% seems to be the # fairest game
expected_value <- mean(replicate(100, game_chain(n=100000, prob_inc=.012)))
expected_value

# This runs the game with a probability increment of 1.3%
expected_value <- mean(replicate(100, game_chain(n=100000, prob_inc=.013)))
expected_value
```

4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
  
```{r}
# Returns a vector of sorted statistics on each sample
bootstrap <- function(data, N, FUN = mean){
  n <- length(data)
  
  # N is how many times we want to sample from our data
  data_sample <- matrix(NA, nrow = N, ncol = n)
  # Store each sample as rows
  for(i in 1:N){
    data_sample[i,] <- sample(data, n, replace = TRUE)
  }
  # Apply specified function by row and sort
  return(sort(apply(data_sample, 1, FUN)))
}

# Test
x <- c(0, 1, 2, 3, 4, 5, 6, 7, 8)
bootstrap(x, 100, FUN=sd)

```
  
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  
```{r}
# Calculating statistic on bootstrap sample taking one sample
bootstrap(x, 1, mean)
bootstrap(x, 1, var)
#bootstrap(x, 1, regression)


```
  
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 

```{r}  
# Calculating statistic on 1000 bootstrap sample 
bootstrap(x, 1000, mean)
bootstrap(x, 1000, sd)
#bootstrap(x, 1, regression)
```  

  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile

Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 

```{r}
# Returns 95% confidence interval of specified statistics on data
boot_ci <- function(data, N, FUN = mean){
  n <- length(data)
  
  # N is how many times we want to sample from our data
  data_sample <- matrix(NA, nrow = N, ncol = n)
  for(i in 1:N){
    data_sample[i,] <- sample(data, n, replace = TRUE)
  }
  # Apply specified function and sort
  functioned_sorted <- (sort(apply(data_sample, 1, FUN)))
  two_five_percentile <- quantile(functioned_sorted, .025)
  ninety_seven_five_percentile <- quantile(functioned_sorted, .975)
  return(c(two_five_percentile, ninety_seven_five_percentile))
}

# Test
boot_ci(x, 1000, mean)
boot_ci(x, 1000, var)
```

5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?